---
title: "PS270 Final Project"
author: "Farrah J. Israel"
description: "My final project"
output:
  html_document
---

# Introduction

My final project sits at the center of what is known as Transitional Justice research which I will be working on with Professor Bates this summer through a summer research grant called L&S Honors Summer Research Apprenticeship. Within Transitional Justice research is this idea of the global "justice cascade" which is the global shift toward holding leaders accountable for human rights abuses. Scholars such as Kathryn Sikkink argue that the modern "justice cascade" has been propelled not only by states but also by human rights organizations. The goal of this overall project is to determine whether atrocities from certain countries and/or certain kinds of atrocities are more/less likely to make its way all the way up to be recognized by large human rights organizations (such as the UN, Amnesty International, and the Red Cross). The data that I am working with for this final project was collected by previous Research Assistants and will serve as the foundation for the work I will be continuing this summer.

This final project specifically, will explore whether a country that has higher democracy levels, has greater access to justice. Access to justice can be defined as the ability of ordinary people to use legal systems to resolve disputes and seek protection. It varies dramatically across countries, and political scientists constantly debate what explains this variation. My research question is: Do higher levels of liberal democracy improve access to justice? I hypothesize that countries with stronger democratic institutions will have higher access to justice scores. This question matters because access to justice is a main component when it comes to rule of law and human rights. In theory, democracies are meant to constrain state power, reduce arbitrary violence, and create institutions for resolving disputes. If democracy actually improves legal access, that has implications for how we think about international development, rule of law promotion, and human rights policy. It also connects to broader patterns visible in the data. Ratifiers of the Rome Statute, for example, tend to be a lot more democratic than countries that do not ratify the Rome Statute.

# Load the data

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
load("RAData.RData")
```

# Rename Columns

```{r}
RAdata <- RAdata |>
  rename(
    Country = country,
    Year = year,
    Intervention = intervention,
    Investigation = investigation,
    Prosecute = prosecute,
    Liberal_Democracy = v2x_libdem,
    Access_to_Justice = v2xcl_acjst,
    Civil_Society = v2xcs_ccsi,
    Atrocities = log_atrocities,
    ICC_Involvement = involvementm,
    Cooperation_Investigation = stateCinv,
    Cooperation_Prosecution = stateCpros)
RAdata
```
# Add Continent Column

```{r}
Africa <- c("Algeria","Angola","Benin","Botswana","Burkina Faso","Burundi","Cameroon",
  "Cape Verde","Central African Republic","Chad","Comoros","Cote d'Ivoire",
  "Democratic Republic of Congo","Egypt","Equatorial Guinea","Eritrea","Ethiopia",
  "Gambia","Ghana","Guinea","Kenya","Lesotho","Liberia","Libya","Madagascar","Malawi",
  "Mali","Mauritania","Mauritius","Morocco","Mozambique","Namibia","Niger","Nigeria",
  "Republic of Congo","Rwanda","Senegal","Sierra Leone","Somalia","South Africa",
  "South Sudan","Sudan","Tanzania","Togo","Tunisia","Uganda","Zambia","Zimbabwe")
North_America <- c("Canada","United States","Mexico","Costa Rica","Cuba",
  "Dominican Republic","El Salvador","Grenada","Guatemala","Haiti","Honduras",
  "Jamaica","Nicaragua","Panama")
South_America <- c("Argentina","Bolivia","Brazil","Chile","Colombia","Ecuador",
  "Paraguay","Peru","Suriname","Uruguay","Venezuela")
Asia <- c("Afghanistan","Azerbaijan","Bahrain","Bangladesh","Bhutan","Cambodia","China",
  "Cyprus","Georgia","Hong Kong","India","Indonesia","Iran","Iraq","Israel","Japan",
  "Jordan","Kazakhstan","Kuwait","Lebanon","Malaysia","Mongolia","Myanmar","Nepal",
  "North Korea","Pakistan", "Philippines","Qatar","Saudi Arabia","Singapore",
  "South Korea","Sri Lanka","Syria","Taiwan","Thailand","Timor-Leste","Turkey",
  "United Arab Emirates","Uzbekistan","Vietnam","Yemen")
Europe <- c("Albania","Austria","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria",
  "Croatia","Czech Republic","Denmark","Finland","France","Germany","Greece","Hungary",
  "Iceland","Ireland","Italy","Kosovo","Latvia","Liechtenstein","Lithuania","Malta",
  "Moldova","Netherlands","Northern Ireland","Norway","Poland","Portugal","Romania",
  "Russia","Serbia","Slovakia","Slovenia","Spain","Sweden","Switzerland","Ukraine",
  "United Kingdom","Yugoslavia")
Oceania <- c("Australia","Fiji","Marshall Islands","Micronesia","New Zealand","Palau","Vanuatu")

RAdata <- RAdata |>
  mutate(
    Continent = case_when(
      Country %in% Africa ~ "Africa",
      Country %in% North_America ~ "North America",
      Country %in% South_America ~ "South America",
      Country %in% Asia ~ "Asia",
      Country %in% Europe ~ "Europe",
      Country %in% Oceania ~ "Oceania",
      TRUE ~ NA_character_),
    P5_label = ifelse(p5 == 1, "P5 Member", "Non-P5"))

continent_colors <- c(
  "Africa" = "#1b9e77",
  "Asia" = "#d95f02",
  "Europe" = "#7570b3",
  "North America" = "#e7298a",
  "South America" = "#66a61e",
  "Oceania" = "#e6ab02")

RAdata
```

# Sort by One Country Per Row

```{r}
RAdata_country_avg <- RAdata |>
  select(-Year) |>
  group_by(Country) |>
  summarize(across(where(is.numeric), mean, na.rm = TRUE))
RAdata_country_avg
```

# Data

This analysis uses country‑year observational data. This makes the research design cross‑sectional with repeated observations, not experimental. The main dependent variable is the Access to Justice Index (v2xcl_acjst). It is a continous measure from 0-1, low to high, where higher values indicate stronger access to justice. The data was calculated and obtained from the Varieties of Democracy (V-Dem) dataset which is based on the expert‑survey question, “Do citizens enjoy secure and effective access to justice?”

The key independent variable is the Liberal Democracy Score (v2x_libdem). It is also a continuous measure from 0-1, low to high, where higher values indicate stronger Liberal Demoracy. The data was also calculated and obtained from the Varieties of Democracy (V-Dem) dataset. For this variable, the data comes from the question: “To what extent is the ideal of liberal democracy achieved?

Additional variables used for descriptive context include continent (categorical), total atrocities (count), and Rome Statute ratification (binary). Atrocities are measured using conflict‑event data that record killings, disappearances, and other forms of political violence.

My hypothesis will be supported if the coefficient on Liberal Democracy is positive and statistically significant. In other words, if more democratic countries systematically score higher on Access to Justice. A coefficient near zero or a p-value above 0.05 would undermine the hypothesis.

To visualize the main outcome variable, the main plot shows the distribution of Access to Justice scores across all country-year observations.

Here is the codebook that goes into more detail about the data: https://drive.google.com/file/d/1rBsOnN3ph3bW2U9ybJmleacqfomuf_k4/view?usp=sharing


## Average Atrocities by Continent

The bar chart below provides descriptive context for why access to justice matters: Africa and Asia show the highest average levels of political violence (measured as logged atrocity counts), suggesting that regions with weaker rule of law face the most severe consequences. This motivates the central analysis, which examines whether stronger democratic institutions are associated with greater access to justice across all regions.

```{r}
RAdata |>
  drop_na(Continent, Atrocities) |>
  group_by(Continent) |>
  summarize(
    avg_atrocities = mean(Atrocities, na.rm = TRUE),
    n = n()) |>
  arrange(desc(avg_atrocities)) |>
  ggplot(aes(x = reorder(Continent, avg_atrocities), 
             y = avg_atrocities, 
             fill = Continent)) +
  geom_col() +
  scale_fill_manual(values = continent_colors) +
  labs(
    title = "Average Atrocities by Continent",
    x = "Continent",
    y = "Average Total Atrocities") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"))
```
```{r}
RAdata |>
  drop_na(Continent, Access_to_Justice) |>
  ggplot(aes(x = reorder(Continent, Access_to_Justice, FUN = median),
             y = Access_to_Justice,
             fill = Continent)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = continent_colors) +
  labs(
    title    = "Distribution of Access to Justice by Continent",
    subtitle = "Each box spans the middle 50% of country-year observations",
    x        = "Continent",
    y        = "Access to Justice Index (0–1)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "none")
```


#Results

## Plot Main Analysis

```{r}
RAdata |>
  filter(!is.na(Liberal_Democracy), !is.na(Access_to_Justice)) |>
  ggplot(aes(x = Liberal_Democracy, y = Access_to_Justice)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title    = "Does Democracy Improve Access to Justice?",
    subtitle = "Country-Year Observations Worldwide",
    x = "Level of Liberal Democracy",
    y = "Access to Justice Index"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

## Another Way to View Plot Main Analysis --- Country Averages
```{r}
RAdata_country_avg |>
  filter(!is.na(Liberal_Democracy), !is.na(Access_to_Justice)) |>
  ggplot(aes(x = Liberal_Democracy, y = Access_to_Justice)) +
  geom_point(alpha = 0.75, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title    = "Does Democracy Improve Access to Justice?",
    subtitle = "Country-Year Observations Worldwide",
    x = "Level of Liberal Democracy",
    y = "Access to Justice Index"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

## Regression

```{r eval}
fit <- lm(Access_to_Justice ~ Liberal_Democracy, data = RAdata)

fit

coef(fit)

library(broom)
glance(fit)

tidy(fit) |>
  select(term, estimate, std.error, p.value) |>
  mutate(across(where(is.numeric), round, 3)) |>
  knitr::kable(
    col.names = c("Variable", "Coefficient", "Std. Error", "P-value"))
```

## Results Write up

The scatterplot Labeled "Does Democracy Improve Access to Justice" has an x-axis of Liberal Democracy Score and a y-axis of Access to Justice Index. This scatterplot has a clear upward trend across all country-year observations. The fitted line goes up to the right, indicating a positive correlation which means that more democratic countries provide greater access to justice. This pattern is pretty consistent across the full distribution of observations. This visual evidence proves my hypothesis correct---democracy and access to justice increase together. 

The regression confirms and also quantifies what the scatterplot shows. The estimated coefficient on Liberal Democracy is 0.841 (Standard Error = 0.009, p < 0.001), meaning there is not only a strong relationship, but also a statistically significant relationship. This means that a one unit increase in the liberal democracy index is associated with an increase of 0.841 on the 0-1 Access to Justice scale. A one-unit increase means moving from fully authoritarian to fully liberal democratic. Since both variables (Liberal Democracy Score and Access to Justice Index) range from 0-1, this coefficient implies that democracy alone accounts for nearly the entire spread in access to justice across counties. The intercept of 0.300 suggests that even countries with the lowest democracy scores are predicted to have some baseline level of access to justice. This means that formal legal institutions exist in authoritarian regimes. The very small p-value (p < 0.001) allows us to reject the hypothesis that liberal democracy has no relationship with access to justice. 

Despite the strength of the association, the results should not be interpreted causally. The research design is observational and cross-sectional with repeated observations, which means that countries are compared to one another rather than randomly assigned different levels of democracy. There are several confounders that could bias the relationship. For example, wealthier countries tend to be more democratic and better able to fund courts, legal aid, and judicial oversight. Another example is the fact that countries with stronger bureaucratic and administrative capacity tend to have ore effective courts and legal institutions and have more stable democratic systems. These would ultimately result in a higher access to justice index. Moreover, because both variables are constructed from the same V-Dem expert-survey methodology, shared measurement processes may artificially inflate the correlation. For those reasons, the coefficient should be interpreted as a strong descriptive association, not a caussal effect of democratization on legal access. 

# Conclusion

Overall, the visualization and regression results provide strong evidence that countries with higher levels of liberal democracy tend to exhibit much greater access to justice. This is clear support for the hypothesis that democratic institutions and legal protections move together. At the same time, however, the analysis is still limited due to the observational nature of the data, the possibility of omitted variables such as economic development or state capacity, and the fact that both variables came from the same expert survey infrastructure (V-Dem). With additional time or resources, the analysis could be strengthened by incorporating a broader set of controls and using alternative data sources to reduce common-method bias. 